---
title: "Practical Machine Learning Course Project Report"
author: "Eduardo Rodríguez"
date: "October 24, 2015"
output: html_document
---

```{r, include=FALSE, cache=FALSE}
library("knitr")
# Set global chunk options
opts_chunk$set(fig.align='center', fig.show='hold')
# Set global hooks
knit_hooks$set(inline = identity)
# Load necessary R packages
library("dplyr")
library("ggplot2")
library("caret")
```

# Introduction

This is the Course Project Report for Practical Machine Learning.

# Model building

We use the RRF ("Regularized Random Forest") method from the caret R package
to fit a random forest model to the data.

```{r, echo=FALSE}
# First, download the data

trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

setwd("/home/eduardo/datascience/coursera/8-predmachlearn/CP/")

trainFile <- "pml-training.csv"
if (!file.exists(trainFile)) {
    download.file(url = trainURL, destfile = trainFile, method = "curl")
}

testFile <- "pml-testing.csv"
if (!file.exists(testFile)) {
    download.file(url = testURL, destfile = testFile, method = "curl")
}

pmlTrain <- read.csv(trainFile, nrows = 19630)
pmlTest <- read.csv(testFile)

# Create training and testing subsets (from the original training data)

inTrain <- createDataPartition(pmlTrain$classe, p = 0.7, list = FALSE)
training <- pmlTrain[ inTrain,]
testing <- pmlTrain[-inTrain,]

# Eliminate useless variables and separate dependent from independent

Xtrain <- select(training, -(X:num_window), -classe)
Ytrain <- select(training, classe)
Xtest <- select(testing, -(X:num_window), -classe)
Ytest <- select(testing, classe)
```

# Cross-validation

Random forest algorithms perform cross validation as an integral part
of the training process, so it is not necessary to add an extra
cross-validation step.

# Out of sample error

# Predictions

